---
title: |
    | Subtype-aware batch correction retains biological signal of integrated
    | breast cancer datasets---Supplementary Methods
author: |
    | Gil Tomás
    | gil.tomas@igmm.ed.ac.uk
output:
    html_document:
        code_folding: hide
        toc: true
        toc_float:
            collapsed: false
        toc_depth: 2
        theme: flatly
        highlight: tango
        df_print: paged
---

```{r options, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      echo = FALSE,
                      cache = FALSE,
                      cache.path = "../cache/composition_",
                      cache.lazy = FALSE,
                      fig.align = "left")
## fig.path = "../cache/figCache/composition_",
## fig.align = 'center')
```

```{r libraries}
library(data.table) # data frame manipulation
library(oligo)      # preprocessing oligonucleotide arrays
library(limma)      # differential expression analysis
library(AIMS)       # implements AIMS classifier (non-parametric version of PAM50)
library(sva)        # implements ComBat
library(ggplot2)    # sophisticated plotting framework
library(ggsignif)   # significance bars for ggplot2
library(irr)        # Cohen's Kappa
library(knitr)      # kable function for tables
library(kableExtra) # format kable tables
library(SNAGEE)     # quality metric based on signal-to-noise ratio
```

```{r variables}
## Directories
rdsDir <- "../out/rds"
csvDir <- "../data/csv"
dsetDir <- "../data/rds"
libDir <- "../lib"
graphsDir <- "../out/pdf"

## Affy Chips
chips <- c("p2", "a")

## Normalisation Methods
normMths <- c("frma", "mas5", "rma")

## Colours
cols <- c("Basal" = "red2",
          "Her2" = "purple",
          "Luminal B" = "cadetblue2",
          "Luminal A" = "dodgerblue4",
          "Normal" = "forestgreen")

## Seed
set.seed(42)
```

```{r source, results = "hide"}
scripts2source <- list.files(libDir)
lapply(scripts2source, function(fl) source(file.path(libDir, fl)))
```

# 1---Dataset Acquisition #

Raw CEL files from ten breast cancer gene expression dataset where downloaded
from [GEO](https://www.ncbi.nlm.nih.gov/geo/) and normalized
with
[fRMA](https://bioconductor.org/packages/release/bioc/html/frma.html),
[RMA](https://www.bioconductor.org/packages/release/bioc/html/oligo.html)
and [MAS5](https://www.bioconductor.org/packages/release/bioc/html/affy.html).
Demographics of each dataset are shown in **Table 1**.

```{r datasets, eval = TRUE, cache = FALSE}
## load datasets table
dsets.dfr <- read.csv(file.path(csvDir, "datasets.csv"),
                      stringsAsFactors = FALSE)
dsets.dtb <- data.table(dsets.dfr)
dsets.dtb[, platform := gsub("_", "", platform)]

## split datasets by chip
p2Dsets <- dsets.dtb[, id[grepl("p2", id)]]
aDsets <- dsets.dtb[, id[!grepl("p2", id)]]
```

``` {r table-1, results = "asis", eval = TRUE, cache = FALSE}
## setkey(dsets.dtb, id)
dsets.dtb[, ':=' (dataset = NULL,
                  from = NULL,
                  notes = NULL)]
setnames(dsets.dtb,
         c("ref", "fracER", "fracHER2"),
         c("GSE", "fracER+", "fracHER2+"))
setcolorder(dsets.dtb,
            c("GSE", "id", "platform", "nSamples", "fracER+", "fracHER2+"))
setorder(dsets.dtb,
         "platform", "fracER+", "fracHER2+")
## sir-p2/GSE17907 has had 4 samples removed because they were not labeled HER2-
## (see bellow)
dsets.dtb[GSE == "GSE17907", nSamples := 33]
kable(dsets.dtb[, -2, with = FALSE],
      format = "html",
      digits = 2,
      caption = "Table 1---Demographics of datsets used in this study.") %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = FALSE)
```

```{r expression-sets, eval = TRUE, cache = TRUE, results = "hide"}
## load expression sets into memory
eSet.lst <- setNames(lapply(normMths, function(normMth) {
    setNames(lapply(dsets.dtb[["id"]], function(dsetId) {
        readRDS(file.path(dsetDir,
                          sprintf("%s-%s-affy.Rds",
                                  dsetId, normMth)))
    }), dsets.dtb[["id"]])
}), normMths)

## trim non her2 ihc positive samples from sir-p2 normalised datasets
trimHer2Neg <- function(eset) eset[, sampleNames(eset)[eset$her2 == 1 & !is.na(eset$her2)]]
eSet.lst[["frma"]][["sir-p2"]] <- trimHer2Neg(eSet.lst[["frma"]][["sir-p2"]])
eSet.lst[["mas5"]][["sir-p2"]] <- trimHer2Neg(eSet.lst[["mas5"]][["sir-p2"]])
eSet.lst[["rma"]][["sir-p2"]] <- trimHer2Neg(eSet.lst[["rma"]][["sir-p2"]])
```

# 2---Microarray Dataset Integration
Microarray dataset integration needs to account for technical, non-biological,
variation across multiple batches of independently acquired datasets.  The goal
of conventional batch correction (BC) is to remove batch effects while retaining
biological variation conveyed by each dataset.  Several methods exist to address
this task.  One popular approach
is [ComBat](https://www.bu.edu/jlab/wp-assets/ComBat/Abstract.html), which
proposes "parametric and nonparametric empirical Bayes frameworks for adjusting
data for batch effects that is robust to outliers in small sample sizes and
performs comparable to existing methods for large samples".  However, most BC
integration strategies do not account for imbalanced subtype composition within
datasets.

This manuscript introduces a novel procedure for integrating expression profile
datasets of known dissimilar composition, called **subtype-aware batch
correction** (SABC).  Molecular subtypes are initially assigned to each sample
in each dataset with a publicly available single sample predictor (SSP).  Then,
batch effects are resolved between datasets on a per-subtype basis.  This
two-layered approach allows for the biological specifics captured by the single
sample predictor of choice to be accounted for during the batch correction step,
and thus carried over into the integrated dataset.  By ignoring distinct subtype
compositions between datasets, conventional BC incorrectly apprehends biological
variation as technical batch effect, and consequently distorts true biological
signal in the integrated dataset.

Breast cancer is widely understood to be subdivided into five intrinsic or
molecular subtypes, which can be assigned by gene expression profile single
sample predictors.  Breast cancer hence provides an ideal case stud y for the
evaluation of SABC.  We used conventional BC and SABC to integrate four breast
cancer datasets hybridized onto the Affymetrix HG-U133a chip and six breast
cancer datasets hybridized onto the HG-U133 Plus2 chip (**Table 1**).

To evaluate the performance of each method, we compared the distributions of
expression values for
the
[205225\_at](https://genecards.weizmann.ac.il/cgi-bin/geneannot/GA_search.pl?keyword_type=probe_set_id&array=HG-U133&target=genecards&keyword=205225_at) and
the
[216836\_s\_at](https://genecards.weizmann.ac.il/cgi-bin/geneannot/GA_search.pl?keyword_type=probe_set_id&array=HG-U133&target=genecards&keyword=216836_s_at) probesets
in each chip, respectively targeting for the *ESR1* and *ERBB2* gene
transcripts.  Estrogen receptor (ESR1) and Erb-B2 Receptor Tyrosine Kinase 2
receptor (ERBB2) status are strong predictors of breast cancer prognosis and are
traditionally assessed by immunohistochemestry (IHC).  Post dataset integration,
the expression values of both genes should remain in line with the biological
signal conveyed by the independent assessment given by IHC for both proteins.
In addition, we compared the agreement between single sample predictor class
assignments for individual samples prior and post integration.  The batch
correction procedure should not interfere with the molecular subtype identity of
each sample, and for that reason higher agreement rates should be indicative of
higher transcriptional fidelity of the integrated dataset.

To guide the integration process with SABC, we used two SSPs implemented in
the [Genefu](https://bioconductor.org/packages/release/bioc/html/genefu.html)
Bioconductor package.  The
first, [sorlie2003](http://www.pnas.org/content/100/14/8418), is based on 534
diagnostic genes and is a five-subtype classifier; the
second,
[desmedts2008](http://clincancerres.aacrjournals.org/content/14/16/5158?ck=nck),
is based on three genes (*ESR1*, *ERBB2* and *AURKA*), and is a three-subtype
classifier.  To assess classifier agreement prior and post integration, and in
order to circumvent the redundancy caused by using the same single sample
predictor to integrate and to validate the integration process, we used the
Genefu implementation of
the [PAM50](http://ascopubs.org/doi/abs/10.1200/JCO.2008.18.1370) single sample
predictor, based on 50 genes and yielding a five-subtype classifier.

```{r pre-predictions, results = "hide", eval = TRUE, cache = TRUE}
preds.dfr <- do.call(rbind,
                     lapply(normMths, function(normMth) {
                         do.call(rbind,
                                 lapply(names(eSet.lst[[normMth]]), function(dset) {
                                     eSet <- eSet.lst[[normMth]][[dset]]
                                     e.mtx <- exprs(eSet) # expression matrix
                                     a.dfr <- fData(eSet) # annotation data frame
                                     dfr <- computePreds(e.mtx, a.dfr, doAIMS = FALSE, doIC10 = FALSE)
                                     dfr[["dset"]] <- dset
                                     dfr[["normMth"]] <- normMth
                                     dfr
                                 }))
                     }))
preds.dtb <- data.table(preds.dfr, key = "sampleName")
## write preds.dtb on disk
## write.csv(preds.dtb,
##           file.path(csvDir, "prior-predictions.csv"))
```

```{r preprocess-integration, eval = TRUE, cache = TRUE, results = "hide"}
## concatenated matrices
cc.lst <- list()
cc.lst[["p2"]] <- lapply(eSet.lst, function(lst) do.call(cbind, lapply(lst[p2Dsets], exprs)))
cc.lst[["a"]] <- lapply(eSet.lst, function(lst) do.call(cbind, lapply(lst[aDsets], exprs)))
## log2 mas5 normalised datasets
cc.lst[["p2"]][["mas5"]] <- log2(cc.lst[["p2"]][["mas5"]])
cc.lst[["a"]][["mas5"]] <- log2(cc.lst[["a"]][["mas5"]])
## factor identifying datasets of origin for concatenated matrices
fac.lst <- lapply(list(p2Dsets, aDsets),
                  function(dsets) {
                      factor(do.call(c,
                                     lapply(dsets,
                                            function(dset.id) {
                                                setNames(rep(dset.id,
                                                             dims(eSet.lst[[1]][[dset.id]])[[2]]),
                                                         ## rep(dsets.dtb[id == dset.id, ref],
                                                         rep(dsets.dtb[id == dset.id, GSE],
                                                             dims(eSet.lst[[1]][[dset.id]])[[2]]))
                                            })))
                  })
names(fac.lst) <- chips
## factor with IHC ER status for datasets hybridised in each chip
erStatus.lst <- list()
erStatus.lst[["a"]] <- ordered(do.call(c, lapply(eSet.lst[[1]][aDsets], function(eSet) {
    pData(eSet)[["er"]]
})),
levels = 0:1,
labels = 0:1)
erStatus.lst[["p2"]] <- ordered(do.call(c, lapply(eSet.lst[[1]][p2Dsets], function(eSet) {
    pData(eSet)[["er"]]
})),
levels = 0:1,
labels = 0:1)
## factor with IHC HER2 status for datasets hybridised in each chip
her2Status.lst <- list()
her2Status.lst[["a"]] <- ordered(do.call(c, lapply(eSet.lst[[1]][aDsets], function(eSet) {
    pData(eSet)[["her2"]]
})),
levels = 0:1,
labels = 0:1)
her2Status.lst[["p2"]] <- ordered(do.call(c, lapply(eSet.lst[[1]][p2Dsets], function(eSet) {
    pData(eSet)[["her2"]]
})),
levels = 0:1,
labels = 0:1)
```

```{r integrate, eval = TRUE, cache = TRUE, results = "hide"}
## integration methods
itg.lst <- list()
## 1. BC
itg.lst[["bc"]] <- lapply(chips, function(chip) {
    chip.lst <- cc.lst[[chip]]
    lapply(chip.lst, function(mtx){
        ComBat(dat = mtx,
               batch = fac.lst[[chip]])
    })
})
names(itg.lst[["bc"]]) <- chips
## 2. SABC
## 2.1 with 3-subtype SSP
itg.lst[["sabc3"]] <- lapply(chips, function(chip) {
    setNames(lapply(names(cc.lst[[chip]]), function(nm) {
        cc.mtx <- cc.lst[[chip]][[nm]]
        doSubtypeCombat(mtx = cc.mtx,
                        d.fac = fac.lst[[chip]],
                        subtypePreds.fac = preds.dtb[normMth == nm][colnames(cc.mtx), subtype.desmedt2008])
    }), names(cc.lst[[chip]]))
})
names(itg.lst[["sabc3"]]) <- chips
## 2.2 with 5-subtype SSP
itg.lst[["sabc5"]] <- lapply(chips, function(chip) {
    setNames(lapply(names(cc.lst[[chip]]), function(nm) {
        cc.mtx <- cc.lst[[chip]][[nm]]
        doSubtypeCombat(mtx = cc.mtx,
                        d.fac = fac.lst[[chip]],
                        subtypePreds.fac = preds.dtb[normMth == nm][colnames(cc.mtx), intrinsic.sorlie2003])
    }), names(cc.lst[[chip]]))
})
names(itg.lst[["sabc5"]]) <- chips
## write itg.lst on disk
## saveRDS(itg.lst,
##         file.path(rdsDir,
##                   "integrated-matrices.Rds"))
```

## 2.1---Distortion of Molecular *ESR1* and *ERBB2* Measurements ##

```{r er-boxplots-preprocess, eval = TRUE, cache = TRUE}
## read itg.lst from disk
## itg.lst <- readRDS(file.path(rdsDir,
##                              "integrated-matrices.Rds"))
## From these, we happen to know that the best probeset to represent ESR1
## expression is 205225_at (courtesy of Andy)
esr1Probe <- "205225_at"    # Andy Sims
erbb2Probe <- "216836_s_at" # Nicholas Moir
## build a data table containing relevant information to plot:
## .sample name
## .dataset
## .chip (a or p2)
## .IHC receptor (er or her2)
## .IHC status for receptor (zero or one)
## .values of expression for corresponding diagnostic probe
## .method of correction (nc, bc, sabc3 or sabc5)
## .normalisation procedure (frma, rma or mas5)
cbn.dfr <- expand.grid(chip = chips,
                       ihc.marker = c("er", "her2"),
                       itg.method = c("nc", "bc", "sabc3", "sabc5"),
                       nrm.method = c("frma", "rma", "mas5"),
                       stringsAsFactors = FALSE)
buildBxpltData <- function(chip,
                           ihc.marker,
                           itg.method,
                           nrm.method) {
    prb <- ifelse(ihc.marker == "er", esr1Probe, erbb2Probe)
    if(ihc.marker == "er") {
        status.lst <- erStatus.lst
    } else {
        status.lst <- her2Status.lst
    }
    if(itg.method == "nc") {
        mtx <- cc.lst[[chip]][[nrm.method]]
    } else {
        mtx <- itg.lst[[itg.method]][[chip]][[nrm.method]]
    }
    data.table(sampleName = colnames(mtx),
               dataset = names(fac.lst[[chip]]),
               chip = chip,
               ihc.marker = ihc.marker,
               ihc.status = status.lst[[chip]],
               expr = mtx[prb, ],
               itg.method = itg.method,
               nrm.method = nrm.method)
}
bxplt.dtb <- do.call(rbind,
                     mapply(buildBxpltData,
                            chip = cbn.dfr[["chip"]],
                            ihc.marker = cbn.dfr[["ihc.marker"]],
                            itg.method = cbn.dfr[["itg.method"]],
                            nrm.method = cbn.dfr[["nrm.method"]],
                            SIMPLIFY = FALSE))
## remove samples for which ihc status is unknown
bxplt.dtb <- bxplt.dtb[!is.na(ihc.status) & !is.na(expr)]
## write bxplt.dtb on disk
## write.csv(bxplt.dtb,
##           file.path(csvDir, "boxplot-data.csv"))
```

We compared the distributions of expression values prior and post dataset
integration for the 205225_at probeset in 840 breast tumours, hybridised onto
the HG-U133a chip, from four datasets with distinct fractions of ER+ samples
(**Table 1** and **Fig.1**).  Regardless of the normalisation method, BC
integration significantly distorts *ESR1* expression measurements in samples
from datasets with extreme fractions of ER+ samples (GSE5327 and GSE17705), to
the point where the two distributions no longer can tell the difference between
ER-- and ER+ samples (**Fig.2**, second column of panels).  In both cases, SABC
integration, whether driven by a 3-subtype SSP (SABC3) or a 5-subtype SSP
(SABC5), succeeds in retaining the biological signal conveyed by the IHC status
in datasets with extreme compositions (**Fig.2**, third and fourth columns of
panels).  Further comparison of ESR1 transcript abundance in these two datasets
prior and post integration reveals that expression values depart significantly
from original measurements when datasets are integrated with BC, yet are
preserved from extreme distortion by SABC integration (**Fig.3**).

```{r er-a-boxplots, fig.height = 6, fig.width = 9, eval = TRUE, fig.cap = 'Figure 1---205225_at probeset measurements hybridised onto the HG-U133a chip broken by dataset prior and post integration.  Integration was done using standard batch correction (BC, with ComBat) and subtype-aware batch correction (SABC3, driven by a three-subtype SSP; and SABC5, driven by a five-subtype SSP).  The distributions are further split by ER status, independently assessed by IHC on fresh frozen specimens.  Raw data was normalised with FRMA, RMA and MAS5.'}
ggplot(bxplt.dtb[chip == "a" &
                 ihc.marker == "er" &
                 !is.na(ihc.status) &
                 ## nrm.method == "frma" &
                 !is.na(expr)],
       aes(x = ordered(dataset,
                       levels = c("GSE5327",
                                  "GSE25065",
                                  "GSE2034",
                                  "GSE17705")),
           y = expr,
           colour = ordered(ihc.status,
                            levels = 0:1,
                            labels = c("ER-", "ER+")))) +
    geom_boxplot() +
    xlab("") +
    ylab("ESR1 (205225_at)") +
    scale_y_continuous(limits = c(2.5, 14)) +
    scale_colour_manual(name = "IHC ER Status",
                        values = c("violetred", "steelblue")) +
    facet_grid(ordered(nrm.method,
                       levels = c ("frma",
                                   "rma",
                                   "mas5"),
                       labels = c("FRMA",
                                  "RMA",
                                  "MAS5")) ~
                   ordered(itg.method,
                           levels = c("nc",
                                      "bc",
                                      "sabc3",
                                      "sabc5"),
                           labels = c("Prior to Integration",
                                      "BC",
                                      "SABC3",
                                      "SABC5")),
               scales = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 14),
          legend.position = "top",
          legend.title = element_text(size = 16),
          legend.text = element_text(size = 16))
## ggsave(file.path(graphsDir, "integrate-er-boxplots-a-frma.pdf"),
##        width = 11,
##        height = 6)
```

```{r er-p2-boxplot, fig.height = 8, fig.width = 10, eval = FALSE}
ggplot(bxplt.dtb[dataset %in% c("GSE16446",
                                "GSE23177") &
                 chip == "p2" &
                 ihc.marker == "er"],
       aes(x = ordered(itg.method,
                       levels = c("nc", "bc", "sabc3", "sabc5"),
                       labels = c("Prior to Integration",
                                  "BC",
                                  "SABC3",
                                  "SABC5")),
           y = expr)) +
    geom_boxplot(outlier.shape = NA) +
    scale_y_continuous(limits = c(5, 18)) +
    xlab("") +
    ylab("ESR1 (205225_at)") +
    geom_signif(comparisons = list(c("Prior to Integration", "BC"),
                                   c("Prior to Integration", "SABC3"),
                                   c("Prior to Integration", "SABC5")),
                y_position = c(15.0,
                               16.2,
                               17.4),
                test = "t.test",
                textsize = 3,
                tip_length = 0) +
    facet_grid(ordered(dataset,
                       levels = c("GSE16446",
                                  "GSE23177"),
                       labels = c("GSE16446 (0% ER+)",
                                  "GSE23177 (100% ER+)")) ~
                   ordered(nrm.method,
                           levels = c ("frma",
                                       "rma",
                                       "mas5"),
                           labels = c("FRMA",
                                      "RMA",
                                      "MAS5"))) +
    ggtitle("HG-U133Plus2") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 14),
          plot.title = element_text(size = 16))
## ggsave(file.path(graphsDir, "comparisons-extreme-datasets-plus2.pdf"),
##        width = 11,
##        height = 5)
```

```{r er-a-extreme-datasets-by-er-status, fig.height = 5, fig.width = 8, eval = TRUE, fig.cap = 'Figure 2---Distributions of probeset 205225_at measurements from datasets GSE5327 (n=58, all ER--) and GSE17705 (n=298, all ER+), taken from **Fig.1**, are compared side by side. See **Fig.1** for details.'}
ggplot(bxplt.dtb[dataset %in% c("GSE5327",
                                "GSE17705") &
                 chip == "a" &
                 ihc.marker == "er"],
       aes(x = ordered(dataset,
                       levels = c("GSE5327",
                                  "GSE17705"),
                       labels = c("GSE5327\n0% ER+",
                                  "GSE17705\n100% ER+")),
           y = expr)) +
    geom_boxplot(outlier.shape = NA) +
    scale_y_continuous(limits = c(5, 18)) +
    xlab("") +
    ylab("ESR1 (205225_at)") +
    geom_signif(comparisons = list(c("GSE5327\n0% ER+",
                                     "GSE17705\n100% ER+")),
                y_position = 16,
                vjust = -.4,
                test = "t.test",
                textsize = 3.6,
                tip_length = 0) +
    facet_grid(ordered(nrm.method,
                       levels = c ("frma",
                                   "rma",
                                   "mas5"),
                       labels = c("FRMA",
                                  "RMA",
                                  "MAS5")) ~
                   ordered(itg.method,
                           levels = c("nc",
                                      "bc",
                                      "sabc3",
                                      "sabc5"),
                           labels = c("Prior to Integration",
                                      "BC",
                                      "SABC3",
                                      "SABC5")),
               scales = "free") +
    ## ggtitle("HG-U133a") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 12),
          plot.title = element_text(size = 16))
## ggsave(file.path(graphsDir, "comparisons-extreme-datasets-a.pdf"),
##        width = 11,
##        height = 5)
```

```{r er-a-extreme-datasets-by-integration, fig.height = 5, fig.width = 7, eval = TRUE, fig.cap = 'Figure 3---Distributions of probeset 205225_at measurements from datasets GSE5327 (n=58, all ER--) and GSE17705 (n=298, all ER+), taken from **Fig.1**, are each compared prior and post dataset integration.  See **Fig.1** for details.'}
ggplot(bxplt.dtb[dataset %in% c("GSE5327",
                                "GSE17705") &
                 chip == "a" &
                 ihc.marker == "er"],
       aes(x = ordered(itg.method,
                       levels = c("nc", "bc", "sabc3", "sabc5"),
                       labels = c("Prior to\nIntegration",
                                  "BC",
                                  "SABC3",
                                  "SABC5")),
           y = expr)) +
    geom_boxplot(outlier.shape = NA) +
    scale_y_continuous(limits = c(5, 18)) +
    xlab("") +
    ylab("ESR1 (205225_at)") +
    geom_signif(comparisons = list(c("Prior to\nIntegration", "BC"),
                                   c("Prior to\nIntegration", "SABC3"),
                                   c("Prior to\nIntegration", "SABC5")),
                y_position = c(15.0,
                               16.2,
                               17.4),
                test = "t.test",
                textsize = 3,
                tip_length = 0) +
    facet_grid(ordered(dataset,
                       levels = c("GSE5327",
                                  "GSE17705"),
                       labels = c("GSE5327 (0% ER+)",
                                  "GSE17705 (100% ER+)")) ~
                   ordered(nrm.method,
                           levels = c ("frma",
                                       "rma",
                                       "mas5"),
                           labels = c("FRMA",
                                      "RMA",
                                      "MAS5"))) +
    ## ggtitle("HG-U133a") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 16),
          strip.text.x = element_text(size = 12),
          strip.text.y = element_text(size = 11),
          plot.title = element_text(size = 16))
## ggsave(file.path(graphsDir, "comparisons-extreme-datasets-a.pdf"),
##        width = 11,
##        height = 5)
```

```{r er-p2-boxplots, fig.height = 6, fig.width = 11, eval = FALSE, fig.cap = 'Figure 2---205225_at probeset measurements hybridised onto the HG-U133Plus2 chip. Refer to **Fig.1** for further details.'}
## read bxplt.dtb from disk
## bxplt.dfr <- read.csv(file.path(csvDir, "boxplot-data.csv"),
##                       row.names = 1)
## bxplt.dtb <- data.table(bxplt.dfr, keep.rownames = FALSE)
ggplot(bxplt.dtb[chip == "p2" &
                 ihc.marker == "er" &
                 !is.na(ihc.status) &
                 ## nrm.method == "frma" &
                 !is.na(expr)],
       aes(x = ordered(dataset,
                       levels = c("GSE16446",
                                  "GSE2109",
                                  "GSE21653",
                                  "GSE5460",
                                  "GSE17907",
                                  "GSE23177")),
           y = expr,
           colour = ordered(ihc.status,
                            levels = 0:1,
                            labels = c("ER-", "ER+")))) +
    geom_boxplot() +
    xlab("") +
    ylab("ESR1 (205225_at)") +
    scale_y_continuous(limits = c(2.5, 14)) +
    scale_colour_manual(name = "IHC ER Status",
                        values = c("violetred", "steelblue")) +
    facet_grid(ordered(nrm.method,
                       levels = c ("frma",
                                   "rma",
                                   "mas5"),
                       labels = c("FRMA",
                                  "RMA",
                                  "MAS5")) ~
                   ordered(itg.method,
                           levels = c("nc",
                                      "bc",
                                      "sabc3",
                                      "sabc5"),
                           labels = c("Prior to Integration",
                                      "BC",
                                      "SABC3",
                                      "SABC5")),
               scales = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 14),
          legend.position = "top",
          legend.title = element_text(size = 16),
          legend.text = element_text(size = 16))
## ggsave(file.path(graphsDir, "integrate-er-boxplots-p2-frma.pdf"),
##        width = 11,
##        height = 6)
```

```{r her2-p2-boxplots, fig.height = 6, fig.width = 10, eval = FALSE, fig.cap = 'Figure 3---216836_s_at probeset measurements hybridised onto the HG-U133Plus2 chip. Refer to **Fig.1** for further details.'}
ggplot(bxplt.dtb[chip == "p2" &
                 ihc.marker == "her2" &
                 !is.na(ihc.status) &
                 ## nrm.method == "frma" &
                 !is.na(expr)],
       aes(x = ordered(dataset,
                       levels = c("GSE23177",
                                  "GSE16446",
                                  "GSE2109",
                                  "GSE21653",
                                  "GSE5460",
                                  "GSE17907")),
           y = expr,
           colour = ordered(ihc.status,
                            levels = 0:1,
                            labels = c("HER2-", "HER2+")))) +
    geom_boxplot() +
    xlab("") +
    ylab("ERBB2 (216836_s_at)") +
    scale_y_continuous(limits = c(2.5, 14)) +
    scale_colour_manual(name = "IHC HER2 Status",
                        values = c("darkorange", "navyblue")) +
    facet_grid(ordered(nrm.method,
                       levels = c ("frma",
                                   "rma",
                                   "mas5"),
                       labels = c("FRMA",
                                  "RMA",
                                  "MAS5")) ~
                   ordered(itg.method,
                           levels = c("nc",
                                      "bc",
                                      "sabc3",
                                      "sabc5"),
                           labels = c("Prior to Integration",
                                      "BC",
                                      "SABC3",
                                      "SABC5")),
               scales = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 14),
          legend.title = element_text(size = 16),
          legend.text = element_text(size = 16))
ggsave(file.path(graphsDir, "integrate-her2-boxplots-p2.pdf"),
       width = 9,
       height = 4)
```

Although less pronounced, a similar trend is observed when comparing the
expression values prior and post- dataset integration for the 216836_s_at
probeset, in 1015 breast tumours hybridised onto the HG-U133Plus2 chip, from six
datasets with distinct proportions of HER2+ samples (**Table 1** and **Fig.4**).
Regardless of the normalisation protocol, The ERBB2 probeset distributions
clearly reflect the IHC HER2 receptor status in the two datasets in our analysis
with extreme HER2 compositions (GSE23177, all HER2--; and GSE17907, all HER2+).
When the five datasets are integrated with BC, this biological signal is erased,
yet preserved (albeit to a lesser extent than in the original datasets), when
integrated with SABC3 and SABC5 (**Fig.4**).

```{r her2-p2-extreme-datasets-by-her-status, fig.height = 5, fig.width = 8, eval = TRUE, fig.cap = 'Figure 4---216836_s_at probeset measurements hybridised onto the HG-U133Plus2 chip are shown for datasets GSE23177 (n=116, all HER2--) and GSE17907 (n=37, all HER2+), in the leftmost column.  These two datasets were integrated with GSE16446, GSE21653, GSE5460 and GSE2109 (**Table 1**), with BC, SABC3 and SABC5.  ERBB2 expression values for the samples in the two datasets with extreme HER2+ compositions are then shown post-integration with each of these methods, broken by normalisation procedure (cf. **Fig.1** for more details).'}
ggplot(bxplt.dtb[dataset %in% c("GSE23177",
                                "GSE17907") &
                 chip == "p2" &
                 ihc.marker == "her2"],
       aes(x = ordered(dataset,
                       levels = c("GSE23177",
                                  "GSE17907"),
                       labels = c("GSE23177\n0% HER2+",
                                  "GSE17907\n100% HER2+")),
           y = expr)) +
    geom_boxplot(outlier.shape = NA) +
    scale_y_continuous(limits = c(7, 18)) +
    xlab("") +
    ylab("ERBB2 (216836_s_at)") +
    geom_signif(comparisons = list(c("GSE23177\n0% HER2+",
                                     "GSE17907\n100% HER2+")),
                y_position = 16,
                vjust = -.4,
                test = "t.test",
                textsize = 3.6,
                tip_length = 0) +
    facet_grid(ordered(nrm.method,
                       levels = c ("frma",
                                   "rma",
                                   "mas5"),
                       labels = c("FRMA",
                                  "RMA",
                                  "MAS5")) ~
                   ordered(itg.method,
                           levels = c("nc",
                                      "bc",
                                      "sabc3",
                                      "sabc5"),
                           labels = c("Prior to Integration",
                                      "BC",
                                      "SABC3",
                                      "SABC5")),
               scales = "free") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 12))
```

The different degree to which subtype-aware batch correction successfully
retains biological signal from these two molecular correlates of breast cancer
biology, in datasets with extreme compositions, could be explained by how well
the classifiers used to drive integration capture the underlying biology of the
ER and HER2 receptors.  Because the clinical HER2 breast cancer phenotype is the
result of a gene amplification, it is possible that gene expression classifiers
are less apt to model binary gene expression distributions (HER2) rather than
continuous ones (ER).  In addition, the top level split highlighted by most
molecular characterisations of breast tumours is lead by ER status, consigning
most SSP derived from molecular data to be particularly sensitive to biological
signal conveyed by this marker.

## 2.2---Single Sample Predictor Agreement ##

Single sample predictor assignments for the Genefu implementation of the PAM50
breast cancer classifier were computed for each of the 1015 samples in GSE2109,
GSE21653, GSE5460, GSE16446, GSE23177 and GSE17907, after FRMA normalisation
(**Table 1**).  We then integrated the six datasets with BC, SABC3 and SABC5,
and computed each sample's PAM50 subtype post-integration.  Comparisons of
subtype assignments prior- and post-integration can be seen on **Fig.5**.

``` {r post-predictions, eval = TRUE, cache = TRUE}
itgMths <- names(itg.lst)
annot.lst <- list()
annot.lst[["p2"]] <- fData(eSet.lst[[1]][p2Dsets][[1]])
annot.lst[["a"]] <- fData(eSet.lst[[1]][aDsets][[1]])
postPreds.dfr <- do.call(rbind,
                         lapply(itgMths, function(itgMth) {
                             do.call(rbind,
                                     lapply(chips, function(chip) {
                                         do.call(rbind,
                                                 lapply(normMths, function(normMth) {
                                                     e.mtx <- itg.lst[[itgMth]][[chip]][[normMth]]
                                                     a.dfr <- annot.lst[[chip]]
                                                     dfr <- computePreds(e.mtx,
                                                                         a.dfr,
                                                                         doAIMS = FALSE,
                                                                         doIC10 = FALSE)
                                                     dfr[["itgMth"]] <- itgMth
                                                     ## dfr[["chip"]] <- chip
                                                     dfr[["normMth"]] <- normMth
                                                     dfr
                                                 }))
                                     }))
                         }))
postPreds.dtb <- data.table(postPreds.dfr, key = "sampleName")
crs.dtb <- unique(preds.dtb[, .(sampleName, dset)])
setkey(crs.dtb, sampleName)
postPreds.dtb <- crs.dtb[postPreds.dtb]
## postPreds.dtb[, priorPost := "post"]
## preds.dtb[, ':='(priorPost = "prior",
##                  itgMth = NA)]
preds.dtb[, itgMth := "nc"]
setcolorder(preds.dtb, names(postPreds.dtb))
priorPostPreds.dtb <- rbind(preds.dtb, postPreds.dtb)
gse.dtb <- dsets.dtb[, .(GSE, id)]
setnames(gse.dtb, c("gse", "dset"))
setkey(gse.dtb, dset)
setkey(priorPostPreds.dtb, dset)
priorPostPreds.dtb <- gse.dtb[priorPostPreds.dtb]
priorPostPreds.dtb[, chip := gsub("(.*)-(.*)", "\\2", dset)]
```

```{r pam50-concordance, eval = TRUE, cache = TRUE}
## frma normalisation and Plus2 chip
pam50.dtb <- dcast.data.table(priorPostPreds.dtb[normMth == "frma" &
                                                 chip == "p2"],
                              sampleName + gse ~ itgMth,
                              value.var = "intrinsic.pam50")
## reorder samples by
## 1. dataset
## 2. subtype
## 3. number of concordant observations
set(pam50.dtb,
    j = "nUnq",
    value = apply(pam50.dtb[,
                            .(bc, nc, sabc3, sabc5)],
                  1,
                  function(x) length(unique(x))))
pam50.dtb[, sampleName := ordered(sampleName,
                                  levels = sampleName[order(gse,
                                                            nc,
                                                            nUnq)])]
pam50.dtb[, gse := ordered(gse,
                           levels = pam50.dtb[,
                                              .N,
                                              by = gse][,
                                                        gse[order(-N)]])]
## melt
pam50Mlt.dtb <- melt(pam50.dtb,
                     id.vars = c("sampleName",
                                 "gse",
                                 "nUnq"))
setnames(pam50Mlt.dtb, "variable", "itgMth")
setnames(pam50Mlt.dtb, "value", "pam50")
pam50Mlt.dtb[, pam50 := ordered(pam50,
                                levels = c("Basal",
                                           "Her2",
                                           "LumA",
                                           "LumB",
                                           "Normal"),
                                labels = c("Basal",
                                           "Her2",
                                           "Luminal A",
                                           "Luminal B",
                                           "Normal"))]
pam50Mlt.dtb[, itgMth := ordered(itgMth,
                                 levels = c("nc", "bc", "sabc3", "sabc5"),
                                 labels = c("Prior to\nIntegration",
                                            "BC",
                                            "SABC3",
                                            "SABC5"))]
```

```{r pam50-concordance-plot, fig.height = 5, fig.width = 9, fig.cap = 'Figure 5---PAM50 subtype assignments for 1015 samples from six datasets (**Table 1**) hybridised on the HG-U133Plus2 chip and normalised with FRMA.  Subtype assignements were computed prior to dataset integration (first row) and post integration with BC, SABC3 and SABC5 (subsequent rows).'}
gseLvls <- pam50Mlt.dtb[, levels(gse)]
gseLbls <- gseLvls
gseLbls[length(gseLbls)] <- sprintf("GSE\n%s", gsub("GSE", "", gseLvls[length(gseLbls)]))
## pam50Mlt.dtb[, gse := ordered(gse,
##                               levels = levels(gse),
##                               labels = gseLbls)]
ggplot(pam50Mlt.dtb, aes(x = sampleName,
                         y = 1,
                         fill = pam50,
                         colour = pam50)) +
  geom_tile() +
  scale_fill_manual(values = cols) +
  scale_colour_manual(values = cols) +
    facet_grid(itgMth ~ ordered(gse,
                                levels = levels(gse),
                                labels = gseLbls),
               scale = "free",
               space = "free") +
  theme_bw() +
  theme(legend.position = "top",
        legend.title = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(size = 6),
        strip.text.y = element_text(size = 8))
```

```{r p2-kvals, eval = TRUE, cache = TRUE}
## functions to compute k-values
getKval <- function(dfr) kappa2(dfr)[["value"]]
k.lst <- list()
k.lst[["all"]] <- sapply(itgMths, function(itgMth) {
    getKval(pam50.dtb[, c("nc", itgMth), with = FALSE])
})
setkey(pam50.dtb, gse)
gses <- pam50.dtb[, levels(gse)]
k.lst <- c(k.lst,
           setNames(lapply(gses, function(gse) {
               sapply(itgMths, function(itgMth) {
                   getKval(pam50.dtb[gse,
                                     c("nc", itgMth),
                                     with = FALSE])
  })
}), gses))
k.dtb <- data.table(do.call(rbind,
                            k.lst),
                    keep.rownames = TRUE)
setnames(k.dtb, "rn", "GSE")
setkey(k.dtb, GSE)
setkey(dsets.dtb, GSE)
k.dtb <- dsets.dtb[, c("GSE",
                       "nSamples",
                       "fracER+",
                       "fracHER2+"),
                   with = FALSE][k.dtb]
setnames(k.dtb,
         c("GSE", "nSamples", "fracER+", "fracHER2+",
           sprintf("Cohen's Kappa: %s",
                   toupper(itgMths))))
k.dtb[GSE == "all", nSamples := k.dtb[, sum(nSamples, na.rm = TRUE)]]
setorder(k.dtb, -nSamples)
```

```{r table-2, results = "asis", eval = TRUE}
kable(k.dtb,
      format = "html",
      digits = 2,
      caption = "Table 2---Interrater agreement of PAM50 subtype assessment prior- and post-dataset integration (see text for details).") %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = FALSE)
```

With the exception of the largest dataset, GSE2109, PAM50 interrater agreement
was always higher with SABC integration than with conventional integration.
Incidentally, the datasets that showed lesser subtype agreement post BC
integration are the ones with most extreme ER and HER2 compositions (GSE16446,
GSE23177 and GSE17907).  For these datasets, SABC integration was able to
increase subtype assignment agreement (with the exception of SABC3 for
GSE23177).

Because sample size can further confound the effect of dataset composition
towards retention of biological signal post-integration, we compared the
PAM50 interrate agreement prior and post-integration for datasets GSE2109,
GSE21653, GSE5460, GSE16446 and GSE23177, sampling in each dataset, 116 samples
with the same IHC ER+ proportions as the original datasets.  Comparisons of
subtype assignements prior and post-integration in this experiment can be seen
in **Fig.6** and **Table 3**.

```{r integrate-proportions, eval = TRUE, cache = TRUE, results = "hide"}
ppGses <- c("GSE2109", "GSE21653", "GSE5460", "GSE16446", "GSE23177")
ppIds <- dsets.dtb[GSE %in% ppGses, id]
## 116 is the minimum number of samples
minSmpls <- min(sapply(eSet.lst[["frma"]][ppIds], dim)["Samples", ])
## retrieve expression sets with the minimum number of samples that respect the
## original fraction of ER+ samples
shrinkEset <- function(eset,
                       nSmpls) {
    fracErPos <- sum(eset$er == 1, na.rm = TRUE) / sum(!is.na(eset$er))
    nSmplsErPos <- round(nSmpls * fracErPos)
    nSmplsErNeg <- nSmpls - nSmplsErPos
    erPosSmpls <- sample(sampleNames(eset)[eset$er == 1 & !is.na(eset$er)], nSmplsErPos)
    erNegSmpls <- sample(sampleNames(eset)[eset$er == 0 & !is.na(eset$er)], nSmplsErNeg)
    eset[, c(erPosSmpls, erNegSmpls)]
}
## only frma normalised data
ppShrunk.lst <- lapply(eSet.lst[["frma"]][ppIds],
                       shrinkEset,
                       nSmpls = minSmpls)
## compute prior predictions
ppPreds.dfr <- do.call(rbind,
                       lapply(names(ppShrunk.lst), function(dset) {
                           eSet <- ppShrunk.lst[[dset]]
                           e.mtx <- exprs(eSet)
                           a.dfr <- fData(eSet)
                           dfr <- computePreds(e.mtx,
                                               a.dfr,
                                               doAIMS = FALSE,
                                               doIC10 = FALSE)
                           dfr[["dset"]] <- dset
                           dfr
}))
## concatenate
ppCc.mtx <- do.call(cbind, lapply(ppShrunk.lst, exprs))
pp.fac <- factor(do.call(c, lapply(names(ppShrunk.lst), function(dset) {
    rep(dset, ncol(ppShrunk.lst[[dset]]))
})))
## integrate
ppItg.lst <- list()
ppItg.lst[["bc"]] <- ComBat(dat = ppCc.mtx,
                            batch = pp.fac)
ppItg.lst[["sabc3"]] <- doSubtypeCombat(mtx = ppCc.mtx,
                                        d.fac = pp.fac,
                                        subtypePreds.fac = ppPreds.dfr$subtype.desmedt2008)
ppItg.lst[["sabc5"]] <- doSubtypeCombat(mtx = ppCc.mtx,
                                        d.fac = pp.fac,
                                        subtypePreds.fac = ppPreds.dfr$intrinsic.sorlie2003)
## compute post predictions
ppPost.dfr <- do.call(rbind,
                      lapply(names(ppItg.lst), function(itgMth) {
                          e.mtx <- ppItg.lst[[itgMth]]
                          a.dfr <- annot.lst[["p2"]]
                          dfr <- computePreds(e.mtx, a.dfr, doAIMS = FALSE, doIC10 = FALSE)
                          dfr[["itgMth"]] <- itgMth
                          dfr
}))
ppPost.dtb <- data.table(ppPost.dfr, key = "sampleName")
ppPost.dtb <- crs.dtb[ppPost.dtb]
ppPreds.dtb <- data.table(ppPreds.dfr)
ppPreds.dtb[, itgMth := "nc"]
setcolorder(ppPreds.dtb, names(ppPost.dtb))
ppPriorPost.dtb <- rbind(ppPreds.dtb, ppPost.dtb)
setkey(ppPriorPost.dtb, dset)
ppPriorPost.dtb <- gse.dtb[ppPriorPost.dtb]
ppPriorPost.dtb[, chip := gsub("(.*)-(.*)", "\\2", dset)]
## pam50
ppPam50.dtb <- dcast.data.table(ppPriorPost.dtb,
                                sampleName + gse ~ itgMth,
                                value.var = "intrinsic.pam50")
set(ppPam50.dtb,
    j = "nUnq",
    value = apply(ppPam50.dtb[,
                            .(bc, nc, sabc3, sabc5)],
                  1,
                  function(x) length(unique(x))))
ppPam50.dtb[, sampleName := ordered(sampleName,
                                    levels = sampleName[order(gse,
                                                              nc,
                                                              nUnq)])]
ppPam50.dtb[, gse := ordered(gse,
                             levels = ppPam50.dtb[,
                                                  .N,
                                                  by = gse][,
                                                            gse[order(-N)]])]
ppPam50Mlt.dtb <- melt(ppPam50.dtb,
                       id.vars = c("sampleName",
                                   "gse",
                                   "nUnq"))
setnames(ppPam50Mlt.dtb, "variable", "itgMth")
setnames(ppPam50Mlt.dtb, "value", "pam50")
ppPam50Mlt.dtb[, pam50 := ordered(pam50,
                                  levels = c("Basal",
                                             "Her2",
                                             "LumA",
                                             "LumB",
                                             "Normal"),
                                  labels = c("Basal",
                                             "Her2",
                                             "Luminal A",
                                             "Luminal B",
                                             "Normal"))]
ppPam50Mlt.dtb[, itgMth := ordered(itgMth,
                                   levels = c("nc", "bc", "sabc3", "sabc5"),
                                   labels = c("Prior to\nIntegration",
                                              "BC",
                                              "SABC3",
                                              "SABC5"))]
```

```{r pam50-concordance-proportions-plot, fig.height = 5, fig.width = 9, fig.cap = 'Figure 6---PAM50 subtype assignments for 580 samples from five datasets (**Table 1**) hybridised on the HG-U133Plus2 chip and normalised with FRMA.  For each dataset, 116 samples were randomly selected in order to respect the original IHC ER+ fraction.  Subtype assignments were computed prior- (top row) and post- (subsequent rows) dataset integration with BC, SABC3 and SABC5.'}
ggplot(ppPam50Mlt.dtb, aes(x = sampleName,
                         y = 1,
                         fill = pam50,
                         colour = pam50)) +
  geom_tile() +
  scale_fill_manual(values = cols) +
  scale_colour_manual(values = cols) +
    facet_grid(itgMth ~ ordered(gse,
                                levels = levels(gse)),
               scale = "free",
               space = "free") +
  theme_bw() +
  theme(legend.position = "top",
        legend.title = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(size = 8),
        strip.text.y = element_text(size = 8))
```

```{r p2-kvals-proportions, eval = TRUE, cache = TRUE}
ppK.lst <- list()
ppK.lst[["all"]] <- sapply(itgMths, function(itgMth) {
    getKval(ppPam50.dtb[, c("nc", itgMth), with = FALSE])
})
setkey(ppPam50.dtb, gse)
gses <- ppPam50.dtb[, levels(gse)]
ppK.lst <- c(ppK.lst,
             setNames(lapply(gses, function(gse) {
                 sapply(itgMths, function(itgMth) {
                     getKval(ppPam50.dtb[gse,
                                         c("nc", itgMth),
                                         with = FALSE])
                 })
             }), gses))
ppK.dtb <- data.table(do.call(rbind,
                              ppK.lst),
                      keep.rownames = TRUE)
setnames(ppK.dtb, "rn", "GSE")
setkey(ppK.dtb, GSE)
setkey(dsets.dtb, GSE)
ppK.dtb <- dsets.dtb[, c("GSE",
                         "nSamples",
                         "fracER+",
                         "fracHER2+"),
                     with = FALSE][ppK.dtb]
setnames(ppK.dtb,
         c("GSE", "nSamples", "fracER+", "fracHER2+",
           sprintf("Cohen's Kappa: %s",
                   toupper(itgMths))))
ppK.dtb[GSE != "all", nSamples := 116]
ppK.dtb[GSE == "all", nSamples := ppK.dtb[, sum(nSamples, na.rm = TRUE)]]
setorder(ppK.dtb, -nSamples)
```

```{r table-3, results = "as.is", eval = TRUE}
kable(ppK.dtb,
      format = "html",
      digits = 2,
      caption =
          "Table 3---Interrater agreement of PAM50 subtype assessment prior- and post-dataset integration, with datasets brought down to the same size (see text for details).") %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = FALSE)
```

```{r dea, eval = FALSE, cache = TRUE}
getDiffExprGns <- function(gse,                # dataset to extract phenotype from
                           nGenes = 1e2,       # number of top genes to retrieve from analysis
                           marker = "grade") { # %in% c("grade", "grade")
    stopifnot(marker %in% c("grade", "er"))
    eset <- eSet.lst[["frma"]][[dsets.dtb[GSE == gse, id]]]
    a.dtb <- data.table(fData(eset),
                        keep.rownames = TRUE,
                        key = "rn")
    if(marker == "grade") {
        smpls2kp <- sampleNames(eset)[eset$grade %in% 2:3]
    } else {
        smpls2kp <- sampleNames(eset)[eset$er %in% 0:1]
    }
    ## prior integration
    esetPrior <- eset[, smpls2kp]
    if(marker == "grade") {
        m.mtx <- model.matrix(~ esetPrior$grade == 2)
    } else {
        m.mtx <- model.matrix(~ esetPrior$er == 0)
    }
    fitPrior <- lmFit(esetPrior,
                      m.mtx)
    fitPrior <- eBayes(fitPrior)
    prior.dtb <- data.table(topTable(fitPrior,
                                     coef = 2,
                                     number = nGenes),
                            keep.rownames = TRUE,
                            key = "rn")
    prior.dtb[, itgMth := "nc"]
    ## post integration
    post.dtb <- do.call(rbind,
                        lapply(names(itg.lst), function(itgMth) {
                            if(marker == "grade") {
                                post.mtx <- itg.lst[[itgMth]][["a"]][["frma"]][, smpls2kp]
                            } else {
                                post.mtx <- itg.lst[[itgMth]][["p2"]][["frma"]][, smpls2kp]
                            }
                            fitPost <- lmFit(post.mtx,
                                             m.mtx)
                            fitPost <- eBayes(fitPost)
                            post.dfr <- topTable(fitPost,
                                                 coef = 2,
                                                 number = nGenes)
                            post.dfr[["itgMth"]] <- itgMth
                            data.table(post.dfr,
                                       keep.rownames = TRUE,
                                       key = "rn")
                        }))
    ## add fData to post.dfr
    post.dtb <- a.dtb[post.dtb]
    setcolorder(prior.dtb, colnames(post.dtb))
    data.table(rbind(prior.dtb, post.dtb))
}
## differential expression analysis datasets
deaGses <- c("GSE25065",
             "GSE2034",
             "GSE17705")
deaERGses <- c("GSE21653",
               "GSE5460",
               "GSE2109")
## compute list of differentially expressed genes
dea.dtb <- do.call(rbind,
                   lapply(deaGses, function(g) {
                       dtb <- getDiffExprGns(g, nGenes = 1e2)
                       dtb[, gse := g]
                       dtb
                   }))
deaER.dtb <- do.call(rbind,
                     lapply(deaERGses, function(g) {
                         dtb <- getDiffExprGns(g,
                                               nGenes = 1e2,
                                               marker = "er")
                         dtb[, gse := g]
                         dtb
}))
## compare number of differentially expressed genes across different integration
## procedures
tab4.dtb <- data.table(GSE = deaGses)
gradeDeaGses.mtx <- sapply(deaGses, function(g)
    table(na.omit(eSet.lst[["frma"]][[dsets.dtb[GSE == g, id]]]$grade)))
tab4.dtb[, ':='(nGrade2 = gradeDeaGses.mtx[2, ],
                nGrade3 = gradeDeaGses.mtx[3, ]) ]
setkey(dsets.dtb, "GSE")
setkey(tab4.dtb, "GSE")
tab4.dtb <- dsets.dtb[, c("GSE", "fracER+", "fracHER2+"), with = FALSE][tab4.dtb]
nGenesTab4.dtb <- dea.dtb[,
                          .(nbGenesDE_BC = length(na.omit(intersect(EntrezGene.ID[itgMth == "nc"], EntrezGene.ID[itgMth == "bc"]))),
                            nbGenesDE_SABC3 = length(na.omit(intersect(EntrezGene.ID[itgMth == "nc"], EntrezGene.ID[itgMth == "sabc3"]))),
                            nbGenesDE_SABC5 = length(na.omit(intersect(EntrezGene.ID[itgMth == "nc"], EntrezGene.ID[itgMth == "sabc5"])))),
                          by = gse]
setnames(nGenesTab4.dtb, "gse", "GSE")
setkey(nGenesTab4.dtb, "GSE")
setkey(tab4.dtb, "GSE")
tab4.dtb <- tab4.dtb[nGenesTab4.dtb]

tab5.dtb <- data.table(GSE = deaERGses)
ERDeaGses.mtx <- sapply(deaERGses, function(g)
    table(na.omit(eSet.lst[["frma"]][[dsets.dtb[GSE == g, id]]]$er)))
tab5.dtb[, ':='(nERNeg = ERDeaGses.mtx[1, ],
                nERPos = ERDeaGses.mtx[2, ]) ]
setkey(dsets.dtb, "GSE")
setkey(tab5.dtb, "GSE")
tab5.dtb <- dsets.dtb[, c("GSE", "fracER+", "fracHER2+"), with = FALSE][tab5.dtb]
nGenesTab5.dtb <- deaER.dtb[,
                          .(nbGenesDE_BC = length(na.omit(intersect(EntrezGene.ID[itgMth == "nc"], EntrezGene.ID[itgMth == "bc"]))),
                            nbGenesDE_SABC3 = length(na.omit(intersect(EntrezGene.ID[itgMth == "nc"], EntrezGene.ID[itgMth == "sabc3"]))),
                            nbGenesDE_SABC5 = length(na.omit(intersect(EntrezGene.ID[itgMth == "nc"], EntrezGene.ID[itgMth == "sabc5"])))),
                          by = gse]
setnames(nGenesTab5.dtb, "gse", "GSE")
setkey(nGenesTab5.dtb, "GSE")
setkey(tab5.dtb, "GSE")
tab5.dtb <- tab5.dtb[nGenesTab5.dtb]

```

```{r table-4, results = "as.is", eval = FALSE, cache = FALSE}
setnames(tab4.dtb,
         c("nbGenesDE_BC",
           "nbGenesDE_SABC3",
           "nbGenesDE_SABC5"),
         c("nbGenesDE: BC",
           "nbGenesDE: SABC3",
           "nbGenesDE: SABC5"))
kable(tab4.dtb,
      format = "html",
      digits = 2,
      caption = "Table 4---For each dataset, a list of the top 100 genes differentially expressed between grade 2 and grade 3 expression profiles was computed.  Shown is the number of genes in common with lists generated from the same comparison, after BC, SABC3 and SABC5 integration.") %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = FALSE)
setnames(tab5.dtb,
         c("nbGenesDE_BC",
           "nbGenesDE_SABC3",
           "nbGenesDE_SABC5"),
         c("nbGenesDE: BC",
           "nbGenesDE: SABC3",
           "nbGenesDE: SABC5"))
kable(tab5.dtb,
      format = "html",
      digits = 2,
      caption = "Table 5---For each dataset, a list of the top 100 genes differentially expressed between ER+ and ER-- expression profiles was computed.  Shown is the number of genes in common with lists generated from the same comparison, after BC, SABC3 and SABC5 integration.") %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = FALSE)
```

```{r snagee, eval = FALSE, cache = TRUE}
snageeWrapper <- function(obj, # object must be either eSet or matrix
                          chip) {
    ## retrieve list with two components:
    ## 1. genes: unique EntrezGene.ID, no NAs. Collapse as needed
    ## 2. data: matrix of logged gene expression values, rownames correspond to genes
    if(class(obj) == "ExpressionSet") {
        eset <- obj
        ez <- fData(eset)$EntrezGene.ID
        e.mtx <- exprs(eset)
    } else {
        e.mtx <- obj
        a.dfr <- annot.lst[[chip]]
        ez <- a.dfr[rownames(e.mtx), "EntrezGene.ID"]
    }
    ## retain only non-NA values
    ez2kp <- !is.na(ez)
    ez <- ez[ez2kp]
    e.mtx <- e.mtx[ez2kp, ]
    ## collapse (using which.max)
    .sums <- rowSums(abs(e.mtx), na.rm = TRUE)
    .id <- tapply(1:nrow(e.mtx),
                  ez,
                  function(x) x[which.max(.sums[x])])
    list(genes = ez[.id],
         data = e.mtx[.id, ])
}

qualPrior.dtb <- do.call(rbind,
                         lapply(names(eSet.lst), function(nrmMth) {
                             do.call(rbind,
                                     lapply(names(eSet.lst[[nrmMth]]), function(dsetNm) {
                                         eset <- eSet.lst[[nrmMth]][[dsetNm]]
                                         chip <- gsub("(.*)-(.*)", "\\2", dsetNm)
                                         data.table(dset = dsetNm,
                                                    chip = chip,
                                                    nrmMth = nrmMth,
                                                    itgMth = "nc",
                                                    qualStudy = qualStudy(snageeWrapper(eset,
                                                                                        chip)))
                                     }))
                         }))

qualPost.dtb <- do.call(rbind,
                        lapply(names(itg.lst), function(itgMth) {
                            do.call(rbind,
                                    lapply(names(itg.lst[[itgMth]]), function(chip) {
                                        do.call(rbind,
                                                lapply(names(itg.lst[[itgMth]][[chip]]), function(nrmMth) {
                                                    itg.mtx <- itg.lst[[itgMth]][[chip]][[nrmMth]]
                                                    do.call(rbind,
                                                            lapply(levels(fac.lst[[chip]]), function(dsetNm) {
                                                                smpls2do <- which(fac.lst[[chip]] == dsetNm)
                                                                data.table(dset = dsetNm,
                                                                           chip = chip,
                                                                           nrmMth = nrmMth,
                                                                           itgMth = itgMth,
                                                                           qualStudy = qualStudy(snageeWrapper(itg.mtx[, smpls2do], chip)))
                                                            }))
                                                }))
                                        }))
                            }))

qual.dtb <- rbind(qualPrior.dtb, qualPost.dtb)
setnames(qual.dtb, "dset", "id")
setkey(qual.dtb, id)
setkey(dsets.dtb, id)
qual.dtb <- dsets.dtb[, .(GSE, id)][qual.dtb]
qual.dtb[, itgMth := ordered(itgMth,
                             levels = c("nc", "bc", "sabc3", "sabc5"),
                             labels = c("Prior to Integration",
                                        "BC",
                                        "SABC3",
                                        "SABC5"))]
qual.dtb[, nrmMth := ordered(nrmMth,
                             levels = normMths,
                             labels = toupper(normMths))]

```

```{r snagee-plot, eval = FALSE, height = 2, width = 8}
ggplot(qual.dtb,
       aes(x = itgMth,
           y = qualStudy)) +
    geom_point() +
    facet_grid(nrmMth ~ GSE) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.title.x = element_blank(),
          axis.text = element_text(size = 8),
          axis.title.y = element_text(size = 10),
          strip.text.x = element_text(size = 7),
          strip.text.y = element_text(size = 10))

```
